Clustering time series based on probability distributions across temporal granularities

### 2021-11-01 main.Rmd
methodology 1st para

<!-- time series clustering and its challenges -->
Time-series clustering is the process of unsupervised partitioning of $n$ time-series data into $k$ ($k<n$) groups such that homogeneous time-series are grouped together based on a certain similarity measure. The time-series features, length of time-series, representation technique, and, of course, the purpose of clustering time-series all influence the suitable similarity measure or distance strategy to a meaningful level. The three primary methods to time series clustering (@liao2005clustering) are algorithms that operate directly with distances or raw data points in the time or frequency domain (distance-based), with features derived from raw data (feature-based), or indirectly with models constructed from raw data (model-based). The efficacy of distance-based techniques is highly dependent on the distance measure utilized. Defining an appropriate distance measure for the raw time series may be a difficult task since it must take into account noise, variable lengths of time series, asynchronous time series, different scales, and missing data. Commonly used Distance-based similarity measures as suggested by a decade review of time series clustering approaches (@Aghabozorgi2015-ct) are Euclidean, Pearson's correlation coefficient and related distances, Dynamic Time Warping, Autocorrelation, Short time series distance, Piecewise regularisation, cross-correlation between time series, or a symmetric version of the Kullback–Liebler distances (@liao2007clustering) but on a vector time series data. Among these alternatives, Euclidean distances have high performance but need the same length of data over the same period, resulting in information loss regardless of whether it is on raw data or a smaller collection of features. DTW works well with time series of different lengths (@corradini2001dynamic), but it is incapable of handling missing observations. Surprisingly, probability distributions, which may reflect the inherent temporal structure of a time series have not been considered in determining time series similarity.

@@

The proposed methodology aim to leverage the intrinsic temporal data structure hidden in time series data. The foundation of our method is unsupervised clustering algorithms based exclusively on the time-series data.The similarity measure is the most essential ingredient of time series clustering. First step is to decide what we mean by similar. The existing work on clustering probability distributions assumes we have an iid sample $f_1(v),\dots,f_n(v)$, where $f_i(v)$ denotes the distribution from observation $i$ over some random variable $v = \{v_t: t = 0, 1, 2, \dots, T-1\}$ observed across $T$ time points. So going back to the smart meter example, $f_i(v)$ is the distribution of customer $i$ and $v$ is electricity demand. In this work, instead of considering the probability distributions of the linear time series, we assume that the measured variables across different categories of any cyclic granularity are from different data generating processes. Hence, we want to be able to cluster distributions of the form $f_{i,A,B \dots, {N_C}}(v)$, where $A, B$ represent the cyclic granularities under consideration such that $A = \{a_j: j=1, 2, \dots J\}$,  $B = \{b_k: k  = 1, 2, \dots K\}$ and so on. We consider individual category of a cyclic granularity ($A$) or combination of categories for interaction of cyclic granularities (for e.g. $A*B$) to have a distribution. For example, let us consider we have two cyclic  granularities of interest, $A = {0, 1, 2, \dots, 23}$ representing hour-of-day and  $B = \{Mon, Tue, Wed, \dots, Sun\}$ representing day-of-week. Each customer $i$ consist of a collection of probability distributions. In case individual granularities ($A$ or $B$) are considered, there are  $J = 24$ distributions of the form $f_{i,j}(v)$ or $K = 7$ distributions of the form $f_{i,k}(v)$ for each customer $i$. In case of interaction, $J*K=168$ distributions of the form $f_{i,j, k}(v)$ could be conceived for each customer $i$. As a result, we need to decide how to measure similarities between these collections of univariate probability distributions. There are multiple ways to measure similarities between time series based on probability distributions, depending on the objective of the problem. This paper focuses on looking at the (dis) similarity between underlying distributions that may have resulted in different patterns across different cyclic temporal granularities, that eventually have resulted in the (dis) similarity between time series. It considers a methodology with two appraoches for  finding distances between time series. Both of these appraoches may be useful in a practical context and, depending on the data set, may or may not propose the same customer classification. The obtained distances could be fed into a clustering algorithm to break large data sets into subgroups that can then be analyzed separately. These clusters may be commonly associated with real-world data segmentation. However, since the data is unlabeled a priori, more information is required to corroborate this. The methodology is explained in the Figure \ref{fig:flowchart} and each element of the pipeline is discussed.

>>>>

<!-- time series clustering and its challenges -->
Time-series clustering is the process of unsupervised partitioning of $n$ time-series data into $k$ ($k<n$) meaningful groups such that homogeneous time-series are grouped together based on a certain similarity measure. The time-series features, length of time-series, representation technique, and, of course, the purpose of clustering time-series all influence the suitable similarity measure or distance strategy to a meaningful level. The three primary methods to time series clustering (@liao2005clustering) are algorithms that operate directly with distances or raw data points in the time or frequency domain (distance-based), with features derived from raw data (feature-based), or indirectly with models constructed from raw data (model-based). The efficacy of distance-based techniques is highly dependent on the distance measure utilized. Defining an appropriate distance measure for the raw time series may be a difficult task since it must take into account noise, variable lengths of time series, asynchronous time series, different scales, and missing data. Commonly used Distance-based similarity measures as suggested by a decade review of time series clustering approaches (@Aghabozorgi2015-ct) are Euclidean, Pearson's correlation coefficient and related distances, Dynamic Time Warping, Autocorrelation, Short time series distance, Piecewise regularisation, cross-correlation between time series, or a symmetric version of the Kullback–Liebler distances (@liao2007clustering) but on a vector time series data. Among these alternatives, Euclidean distances have high performance but need the same length of data over the same period, resulting in information loss regardless of whether it is on raw data or a smaller collection of features. DTW works well with time series of different lengths (@corradini2001dynamic), but it is incapable of handling missing observations. Surprisingly, probability distributions, which may reflect the inherent temporal structure of a time series have not been considered in determining time series similarity.

@@

The foundation of our method is unsupervised clustering algorithms based exclusively on the time-series data. The proposed methodology aims to leverage the intrinsic data structure hidden within cyclic temporal granularities. The existing work on clustering probability distributions assumes we have an iid sample $f_1(v),\dots,f_n(v)$, where $f_i(v)$ denotes the distribution from observation $i$ over some random variable $v = \{v_t: t = 0, 1, 2, \dots, T-1\}$ observed across $T$ time points. <!--So going back to the smart meter example, $f_i(v)$ is the distribution of customer $i$ and $v$ is electricity demand.--> In this work, instead of considering the probability distributions of the linear time series, we assume it across different categories of any cyclic granularity. We can consider categories of an individual cyclic granularity ($A$) or combination of categories for two interacting granularities ($A*B$) to have a distribution, where  $A, B$ represent two cyclic granularities defined by $A = \{a_j: j=1, 2, \dots J\}$ and  $B = \{b_k: k  = 1, 2, \dots K\}$. For example, let us consider two cyclic  granularities $hod$ and $dow$ representing hour-of-day and day-of-week and defined by, $hod = \{0, 1, 2, \dots, 23\}$ and $dow = \{Mon, Tue, Wed, \dots, Sun\}$. In case individual granularities ($A$ or $B$) are considered, there are  $J = 24$ distributions of the form $f_{i,j}(v)$ or $K = 7$ distributions of the form $f_{i,k}(v)$ for each customer $i$. In case of interaction, $J*K=168$ distributions of the form $f_{i,j, k}(v)$ could be conceived for each customer $i$. Hence clustering these customers is equivalent to clustering these collections of conditional distributions. Towards this goal, we need to decide how to measure similarities between collections of univariate probability distributions. There are multiple ways to measure similarities depending on the aim of the analysis. <!--This paper focuses on looking at the (dis) similarity between underlying distributions that may have resulted in different patterns across different cyclic temporal granularities, that eventually have resulted in the (dis) similarity between time series. --> This paper considers a methodology with two approaches for finding distances between time series. Both of these approaches may be useful in a practical context and, depending on the data set, may or may not propose the same customer classification. The obtained distances could be fed into a clustering algorithm to break large data sets into subgroups that can then be analyzed separately. <!--These clusters may be commonly associated with real-world data segmentation. However, since the data is unlabeled a priori, more information is required to corroborate this.--> The methodology is explained in the Figure \ref{fig:flowchart} and each element of the pipeline is discussed.
